---
title: "Foundations"
subtitle: "Session 7: A Big LMM"
date: 2025-07-31
# For author options see : 
# https://quarto.org/docs/authoring/front-matter.html#authors-and-affiliations
# NB: multiple authors can be added here.
author:
  - name:
      given: Joshua
      family: Wilson Black
    email: joshua.black@canterbury.ac.nz
    orcid: 0000-0002-8272-5763
    affiliation: 
      - "Te Kāhui Roro Reo | New Zealand Institute of Language, Brain and Behaviour"
      - "Te Whare Wānanga o Waitaha | University of Canterbury"
format:
  revealjs:
    theme: [custom.scss]
    incremental: true
    logo: images/NZILBB-small.svg
    df-print: paged
    template-partials:
      - title-slide.html
    title-slide-attributes:
      # If you have Marsden funding, change image name to `nzilbb-uc-marsden.svg`
      data-background-image: images/nzilbb-uc.svg
      # First number controls the horizontal position, second controls vertical.
      data-background-position: '50% 5%'
      # Controls size of image relative to width of the slide.
      data-background-size: 50%
    embed-resources: false
    include-in-header:
      - text: |
          <link rel="icon" type="image/png" sizes="32x32" href=".//images/fav.png" />
bibliography: 
  - grateful-refs.bib
  - stat_workshops.bib
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
knitr:
  opts_chunk: 
    dev: "png"
    dev.args: 
      bg: "transparent"
---

```{r}
#| echo: false
library(tidyverse)
library(broom.mixed)
library(glue)
library(here)
library(lme4)
library(ggeffects)
library(lattice) # required for some plots in lme4.
library(patchwork)

big_dia <- read_csv(here('data', 'big_dia.csv'))

big_dia <- big_dia |> 
  mutate(
    final = factor(final),
    initial = factor(initial),
    repeated.20 = factor(repeated.20)
  )

theme_set(theme_minimal())
```

## Last Time

- What is a **mixed effects model**?
- Adding random effects to a linear model:
    - Use of `lme4` and the `lmer()` function
    - Random intercepts (`(1|group)`)
    - Random intercepts and slopes (`(1 + variable|group)`)
- Conditional vs. marginal predictions

## This Time

- A complex real world example
    - A Big LMM from [@soskuthyChangingWordUsage2017]
- Some lessons along with way:
    - Extracting random effects
    - More plotting hints


# Real life example

## Change in usage factors

- Our models with data from [@soskuthyChangingWordUsage2017] look at the 
relationship between usage factors and word duration.
- But what about **change** in usage factors.
    - e.g.: if a word becomes more frequent does it also become shorter?
    - e.g.: if a word is more frequently utterance final, does it become shorter
    everywhere?
- Models in the paper address this question.

## {.smaller}

::::: {.columns}

:::: {.column width="70%"}
::: fragment

![Fig. 3(c), [@soskuthyChangingWordUsage2017]](images/panel_c.png){height=650}

:::

::::

:::: {.column width="30%"}

- Change in frequency ($x$ axis) leads to reduction in word duration ($y$).
- Colours: baseline duration of word.
- Points: raw data.
- Dashed lines in main panel: predictions from alternative model.
- Inserts: Duration reductions for word with given change in frequency.

::::
:::::

## Modelling strategies

1. Two-stage model.
    1. Fit a model to control variation due to 'things we want to ignore', 
    2. Extract random effects which capture variation of interest, fit
    a model to these.
2. Single-stage model.
    - Fit a model with both control variables and variables of interest.
- Lots of projects at NZILBB use something like (1).

# 'Control Model'

## Stage 1 (the 'control model')

:::: {.r-stack}

::: {.fragment .fade-in}

```{r}
#| eval: false
control_full <- lmer(
  WordDuration ~ scale(YOB) +
    # Duration controls
    scale(dur.context) + 
    scale(seg.no) +
    scale(syll.no) +
    scale(syll.length.3) +
    # Prosodic controls
    final +
    initial +
    scale(final.prop.log) +
    # Predictability controls
    scale(prev_pred_wf_log) +
    scale(foll_pred_wf_log) +
    scale(prev_info_wf_avg) +
    scale(foll_info_wf_avg) +
    scale(unigram.google.gb) +
    # Repetition controls
    repeated.20 +
    # Word class control
    wclass +
    # Random effects
    (1 + scale(YOB) | TargetOrthography) +
    (1 + scale(prev_info_wf_avg) + scale(foll_info_wf_avg) + 
       scale(final.prop.log) + scale(unigram.google.gb) || Speaker) +
    (1 | Corpus),
    data=big_dia,
    REML=TRUE # changed to fit quickly.
)

# save the model
write_rds(
  control_full,
  here('models', 'control.rds'),
  compress = "gz"
)
```

```{r}
#| include: false
control_full <- read_rds(here('models', 'control.rds'))
```


:::

::: {.fragment .fade-in-then-out .big}
😨
:::

::: {.fragment .fade-in-then-out .big}
🧐
:::

::: {.fragment .fade-in .big .backlight}
- Use of `scale()` in formula (OK)
- Note `||` in random smooths
- Save things which take ages
:::


::::

## Summary (Fixed Effects) {.smaller}

::: fragment
```{r}
#| output-location: fragment
tidy(control_full, effects = "fixed") |> 
  select(-effect)
```

:::

::: r-stack

:::: {.fragment .fade-in-then-out}
Intercept represents duration (s) for an adjective which is not utterance
final or initial and not repeated within 20 seconds, at average value of continuous predictors.
::::

:::: {.fragment .fade-in-then-out}
Strongest predictors are `seg_no` and `syll.length.3`, i.e., how many
segments are in the word and average syllable duration. Long words are long! This is the kind of stuff which we want to _control_.
::::

:::: {.fragment .fade-in-then-out}
Yikes, `finalTRUE` has a negative estimate! Utterance final words are _shorter_ now? This is, due to **collinearity** (more on this later).
::::

:::

## Summary (Random Effects) {.smaller}

```{r}
#| output-location: fragment
tidy(control_full, effects = "ran_pars") |> 
  select(-effect)
```

::: r-stack

:::: {.fragment .fade-in-then-out}
Random effects give us `sd_` values (estimates of variation in the population).
::::

:::: {.fragment .fade-in-then-out}
We also get `cor_` terms for correlation between intercepts and smooths.
Let's visualise what this means:
::::

:::

## Correlation of intercept and slope {.smaller}

```{r}
#| echo: false
#| cache: true
#| fig.cap: |
#|   Correlation of intercept terms and slope terms for distinct words.
word_effects <-ranef(control_full)$TargetOrthography |> 
  as_tibble(rownames = "word")

word_effects <- word_effects |> 
      mutate(
        highlight = `(Intercept)` < -0.1 | 
          (`(Intercept)` > 0.1 & 
             between(`scale(YOB)`, -0.02, -0.01))
      )

word_effects |> 
  ggplot(
    aes(
      x = `(Intercept)`,
      y = `scale(YOB)`
    )
  ) +
  geom_point(aes(size=highlight)) +
  geom_smooth(method="lm") +
  geom_label(
    aes(label = word), 
    data = word_effects |> 
      filter(
        `(Intercept)` < -0.1 | 
          (`(Intercept)` > 0.1 & 
             between(`scale(YOB)`, -0.02, -0.01))
      ),
    nudge_y = -0.004,
    nudge_x = 0.006
  ) +
  theme(
    legend.position = "none"
  )
```

::: r-stack

:::: {.fragment .fade-in-then-out}

'Minute' has a low intercept. It is shorter than expected given fixed effects.

::::

:::: {.fragment .fade-in-then-out}
It also has a slope value 0.01 larger than average for year of birth. That is, it has a slope of `r -0.013 + 0.01`. It changes _less_ than other words over time.
::::

:::: {.fragment .fade-in-then-out}
"Education" is longer than expected (intercept), and has a steeper than expected slope. It changes _more_ than other words.
::::

:::: {.fragment .fade-in-then-out}
The correlation means that words which are _longer_ than expected change _more_ over time.
::::

:::

## What does the model say?

- Let's look at some predictions.
- The `ggeffects` package is an easy option.
- We'll use the `predict_response()` function.
- The default gives us a _conditional_ prediction and tells us
about the conditions it uses.
- These can take a while to generate for large models (like this one!)

## Conditional Prediction

::: fragment
```{r}
#| output-location: slide
#| cache: true
yob_conditional <- predict_response(
  control_full,
  terms = "YOB [1870:1980, by = 10]" # What terms do I want predictions for?
)
plot(yob_conditional)
```
::: 

- Looking at `yob_conditional` will tell you what values 
the prediction is for, incl. e.g., `dur.context = 328.77`

## Marginal predictions

```{r}
#| output-location: slide
#| cache: true
yob_marginal <- predict_response(
  control_full,
  terms = "YOB [1870:1980, by = 10]",
  margin = "empirical"
)
plot(yob_marginal)
```

## Multiple predictors

```{r}
#| output-location: slide
#| cache: true
yob_wcalss_conditional <- predict_response(
  control_full,
  terms = c("YOB [1870:1980, by = 30]", "wclass")
)
plot(yob_wcalss_conditional)
```



## Exercise 1 {background-color="#191818"}

- Download the control model (<https://osf.io/q5wgh/>)
- Put it in a directory called `models` inside your project.
- Load the model using `read_rds()` (make sure you have  `library(tidyverse)` earlier in your script).
- Find the variable names by looking at the model summary.
- Generate and plot a prediction `predict_response()` and
`plot()` from the `ggeffects` package.

## Exercise 1 (2) {background-color="#191818"}

```{r}
#| eval: false
#| message: false
control_mod <- read_rds('models/control_full_ML.rds')
my_prediction <- predict_response(
  control_mod,
  terms = c("dur.context [100:400, by = 50]")
)
plot(my_prediction)
```

## Aside: Collinearity {.smaller}

- What if two variables give the 'same' information to the model?
    - i.e., what if they are correlated?
- The model has nothing to choose between the two.
- Consequence: including highly correlated variables leads to unstable 
coefficients.
    - ... variable 1 might get a big coefficient, sometimes variable 2, sometimes both might get a moderate coefficient.
- In this model: `final` and `dur.context` are correlated.

## Aside: Collinearity (2)

::: fragment
```{r}
#| output-location: slide
big_dia |> 
  ggplot(
    aes(
      x = final,
      y = dur.context
    )
  ) +
  geom_boxplot()
```

:::

## Aside: Collinearity (3)

- If we exclude `dur.context`, then the coefficient for `final` has a more
sensible value.
    - ... i.e. utterance final words are estimated to be _longer_.
- This is reported in the paper.
- This model passes a standard check for collinearity 
(the 'variance inflation factor'). Just because you pass a test, doesn't 
mean there isn't a problem!

# 'Treatment Model'

## Extracting random effects {.smaller}

- Sometimes random effects capture the variation we care about.
- e.g. random effects for **word** include an intercept and slope.
- If we are interested in variation in the **change in duration over time**, the slopes capture what we are after.
    - i.e. we want the bit of the model corresponding to `(1 + scale(YOB) | TargetOrthography)`.
- As previously, use the `ranef()` function.

## From earlier

```{r}
word_effects |> 
  slice_sample(n=8)
```

## 🙈🙉🙊

::: fragment
```{r}
words <- word_effects |> 
  left_join(
    big_dia |> 
      select(
        word = TargetOrthography, contains('info'),
        contains('diff'), 'dur.context.avg', 'unigram.google.gb',
        'final.prop.log'
      ) |> 
      unique()
  ) |> 
  mutate(
    # i.e. change in seconds over 100 years
    dur.change = `scale(YOB)` * (100/sd(big_dia$YOB))
  )

# Scale with respect to original data set (to compare with control model)
for (var in c("foll_info_wf_avg", "prev_info_wf_avg", "final.prop.log", 
              "foll.diff", "prev.diff", "final.diff", "unigram.google.gb", 
              "unigram.diff.google.gb", "dur.context.avg")) {
  words[,paste(var, ".s", sep="")] <- (
    (words[,var] - mean(as.numeric(big_dia[[var]]))) / 
      sd(as.numeric(big_dia[[var]]))
  )
}
```

:::

## The Treatment Model

::: {.fragment .fade-in}

```{r}
#| output-location: slide
treatment_model <- lm(
  dur.change ~ (foll_info_wf_avg.s + 
    prev_info_wf_avg.s + 
    final.prop.log.s +
    foll.diff.s + 
    prev.diff.s +
    final.diff.s + 
    unigram.google.gb.s +
    unigram.diff.google.gb.s)*dur.context.avg.s,
  data = words
)
tidy(treatment_model) |> 
  filter(p.value < 0.05)
```

:::

:::: {.r-stack}

::: {.fragment .fade-in-then-out}
This is a multiple linear regression not a mixed model. Why?
:::

::: {.fragment .fade-in-then-out}
Note `(var_1 + var_2) * var_3` syntax. Short for `var_1 * var_3 + var_2 * var_3`. 
:::


::::

## Plot predictions

```{r}
#| output-location: slide
freq_dur_predictions <- predict_response(
  treatment_model,
  terms = c("unigram.diff.google.gb.s", "dur.context.avg.s"),
  margin = "empirical"
)
plot(freq_dur_predictions)
```

## Plot predictions (2)

```{r}
#| output-location: slide
final_dur_predictions <- predict_response(
  treatment_model,
  terms = c("final.diff.s", "dur.context.avg.s"),
  margin = "empirical"
)
plot(final_dur_predictions)
```

## Plot predictions (3)

```{r}
#| output-location: slide
prev_predictions <- predict_response(
  treatment_model,
  terms = c("prev.diff.s"),
  margin = "empirical",
)
plot(prev_predictions) +
  geom_point(
    aes(
      x = prev.diff.s, y = dur.change
    ), 
    data = words,
    alpha = 0.4
  )
```

## Upshot {.smaller}

> [Words that were increasing/decreasing in informativity (based on the previous
contexts) showed a change in the same direction in duration (e.g. increasing
informativity is associated with increasing duration).]{.fragment} [Long words
that were becoming more frequent were also becoming shorter.]{.fragment} [Short
words that were increasingly appearing utterance finally were also becoming
longer.]{.fragment} [It is important to recall that the treatment model holds
constant the local effects of the predictors. Thus, the results show – for
example – that words that are increasing in utterance finality are also
increasing in duration, even when the local position of each token is accounted
for.]{.fragment}

## Exercise 2 {background-color="#191818"}

- Extract the by-speaker random effects.
    - `ranef(control_full)$???`
- Make a plot using this data.
    - e.g. histogram of by-speaker random intercepts.
- Extract the by-word effects (following slides) and 
fit the treatment model. 
- Plot another prediction from the model.

## Next Time

- Preregister a prediction about word duration in QuakeBox
- Run the analysis
- Write it up

# References

```{r}
#| echo: false
grateful::nocite_references(
  grateful::cite_packages(output = "citekeys", out.dir = here())
)
```


::: refs

:::
